---
title: "World Development Indicators (2022) Analysis"
author: "AC"
format: revealjs
editor: visual
theme: moon
transition: slide
---

### World Development Indicators Analysis

Course: QTM 350

*Date: October 6, 2025*

## Intro

-   Analysis of 2022 World Development Indicators

-   Indicators: Education expenditure, health expenditure, life expectancy

## Key Findings on Education Expenditure

-   Range: 0.3% to 16.58% of GDP

-   Top 10: Kiribati leads, followed by small/developing nations

```{r}
#| label: fig-edu-top10-slide
#| echo: false
#| fig-cap: "Top 10 countries by education expenditure as a share of GDP (2022). Source: World Development Indicators (wdi.csv)."
#| fig-width: 8
#| fig-height: 5
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
wdi <- read_csv("wdi.csv", na = "")


top10_edu <- wdi %>%
  filter(!is.na(education_expenditure_gdp_share)) %>%
  slice_max(order_by = education_expenditure_gdp_share, n = 10) %>%
  arrange(education_expenditure_gdp_share)

ggplot(top10_edu,
       aes(x = reorder(country, education_expenditure_gdp_share),
           y = education_expenditure_gdp_share)) +
  geom_col() +
  coord_flip() +
  labs(title = "Education Expenditure (% of GDP), Top 10 Countries (2022)",
       x = NULL,
       y = "Education expenditure (% of GDP)")
```

## Health Expenditure vs. Life Expectancy

-   Range: Health expenditure 1.2% to 16.57% (U.S. outlier), Life expectancy 53.00 to 85.38 years.

    ```{r}
    #| label: fig-health-life-slide
    #| echo: false
    #| fig-cap: "Health expenditure (% of GDP) vs. life expectancy (years), 2022. Line shows linear fit. Source: World Development Indicators (wdi.csv)."
    #| fig-width: 7
    #| fig-height: 5
    ggplot(wdi, aes(x = health_expenditure_gdp_share, y = life_expectancy)) +
      geom_point(na.rm = TRUE) +
      geom_smooth(method = "lm", se = FALSE, na.rm = TRUE) +
      labs(title = "Does Higher Health Spending Share Relate to Longevity?",
           x = "Health expenditure (% of GDP)",
           y = "Life expectancy (years)")
    ```

## Summary

Education spending varies widely, with developing nations leading; health spending shows a weak link to life expectancy.

Future: Analyze regional differences and health system efficiency.\
