---
title: "World Development Indicators (2022) Report"
author: "AC"
date: "October 6, 2025"
format: 
  revealjs:
    output-dir: docs
    theme: dracula
    transition: fade
    self-contained: true
    zoom: 0.65
execute:
  echo: false
  warning: false
---

## Intro

-   The World Development Indicators (WDI) dataset, compiled by the World Bank, provides a comprehensive collection of internationally comparable statistics about global development (World Bank, 2022).

-   Highlight three core indicators:

    -   Education expenditure (% of GDP)
    -   Health expenditure (% of GDP)
    -   Life expectancy (in years)

-   Analyzes resource allocation and human capital development.

## Key Findings on Education Expenditure

```{r}
#| label: fig-edu-top10-slide
#| echo: false
#| fig-cap: "Top 10 countries by education expenditure as a share of GDP (2022). Source: World Development Indicators (wdi.csv)."
#| fig-width: 8
#| fig-height: 5
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
wdi <- read_csv("wdi.csv", na = "")


top10_edu <- wdi %>%
  filter(!is.na(education_expenditure_gdp_share)) %>%
  slice_max(order_by = education_expenditure_gdp_share, n = 10) %>%
  arrange(education_expenditure_gdp_share)

ggplot(top10_edu,
       aes(x = reorder(country, education_expenditure_gdp_share),
           y = education_expenditure_gdp_share)) +
  geom_col() +
  coord_flip() +
  labs(title = "Education Expenditure (% of GDP), Top 10 Countries (2022)",
       x = NULL,
       y = "Education expenditure (% of GDP)")
```

Kiribati leads, suggesting small nations prioritize education.

## Health Expenditure vs. Life Expectancy

```{r}
#| label: fig-health-life-slide
#| echo: false
#| fig-cap: "Health expenditure (% of GDP) vs. life expectancy (years), 2022. Line shows linear fit. Source: World Development Indicators (wdi.csv)."
#| fig-width: 7
#| fig-height: 5
ggplot(wdi, aes(x = health_expenditure_gdp_share, y = life_expectancy)) +
  geom_point(na.rm = TRUE) +
  geom_smooth(method = "lm", se = FALSE, na.rm = TRUE) +
  labs(title = "Does Higher Health Spending Share Relate to Longevity?",
       x = "Health expenditure (% of GDP)",
       y = "Life expectancy (years)")
```

Weak positive association, limited by 197 missing values.

## Summary Table

```{r}
#| label: tbl-key-stats
#| echo: false
#| tbl-cap: "Key statistics for 2022 indicators. Source: World Development Indicators (wdi.csv)."

key_stats <- data.frame(
  Indicator = c("Education Expenditure (% GDP)", "Health Expenditure (% GDP)", "Life Expectancy (Years)"),
  Mean = c(mean(wdi$education_expenditure_gdp_share, na.rm = TRUE),
           mean(wdi$health_expenditure_gdp_share, na.rm = TRUE),
           mean(wdi$life_expectancy, na.rm = TRUE)),
  Median = c(median(wdi$education_expenditure_gdp_share, na.rm = TRUE),
             median(wdi$health_expenditure_gdp_share, na.rm = TRUE),
             median(wdi$life_expectancy, na.rm = TRUE)),
  Min = c(min(wdi$education_expenditure_gdp_share, na.rm = TRUE),
          min(wdi$health_expenditure_gdp_share, na.rm = TRUE),
          min(wdi$life_expectancy, na.rm = TRUE)),
  Max = c(max(wdi$education_expenditure_gdp_share, na.rm = TRUE),
          max(wdi$health_expenditure_gdp_share, na.rm = TRUE),
          max(wdi$life_expectancy, na.rm = TRUE)),
  Missing = c(sum(is.na(wdi$education_expenditure_gdp_share)),
              sum(is.na(wdi$health_expenditure_gdp_share)),
              sum(is.na(wdi$life_expectancy)))
)
knitr::kable(key_stats, align = "c", digits = 2)
```

Wide ranges and missing data suggest reporting gaps.

## Insights

-   Top ten education spenders are predominantly small or developing nations, indicating that percentage-of-GDP metrics can overstate investment intensity.
-   The weak but positive relationship between health expenditure and life expectancy supports the view that financial inputs are necessary but insufficient; system design and efficiency are crucial.
-   The high level of missing data highlights an urgent need for improved international reporting standards.
-   Modest correlation between education expenditure and life expectancy reflects broader structural factors.

## Future Implications

Policymakers could focus on:

-   enhancing efficiency and quality of investments

-   prioritizing early childhood education and preventive healthcare

-   improving data transparency.

## References

-   Center for Sustainable Systems, University of Michigan. (2025). *Social development indicators factsheet*. https://css.umich.edu/publications/factsheets/sustainability-indicators/social-development-indicators-factsheet
-   Harry A. Patrinos. (2023). *Investing in tomorrow*. https://blogs.worldbank.org/en/education/investing-tomorrow-how-educational-spending-translates-lifelong-returns
-   Learning Policy Institute. (2024). *How money matters*. https://learningpolicyinstitute.org/product/how-money-matters-factsheet
-   World Bank. (2022). *World development indicators*. https://databank.worldbank.org/source/world-development-indicators

## Thank you!
